---
title: "Review of Statistics and Probability III"
subtitle: "HS 821 · Lecture 3 · Topics in Time Series Econometrics"
author: "Sunil Paul"
date: "2026-08-12"
format:
  revealjs:
    theme: default
    slide-number: c/t
    show-slide-number: all
    incremental: false
    smaller: false
    scrollable: true
    center-title-slide: true
    transition: slide
    background-transition: fade
    chalkboard: true
    code-line-numbers: false
    code-overflow: wrap
    html-math-method: mathjax
    fig-align: center
    footer: "HS 821 · Review of Statistics & Probability III"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dev: "png"
    dpi: 150
    fig.width: 7
    fig.height: 4.2
---

```{=html}
<style>
:root{
  --hs-accent:#1b3a6b;
  --hs-accent2:#c0392b;
}
.reveal h1,.reveal h2,.reveal h3{ color:var(--hs-accent); font-weight:650; }
.reveal .slide-number{ background:transparent; color:#888; }
.reveal section img{ border:none; box-shadow:none; }
.reveal table{ font-size:0.80em; }
.reveal table th{ background:#eef2f8; }
.reveal pre code{ font-size:0.72em; line-height:1.25; }
.reveal .small{ font-size:0.78em; }
.reveal .muted{ color:#5a6472; }
.reveal blockquote{
  border-left:4px solid var(--hs-accent);
  background:#f6f8fc; padding:0.4em 0.9em; font-style:normal;
}
.callout-ts{
  border-left:5px solid var(--hs-accent2);
  background:#fdf1ef; padding:0.5em 0.9em; border-radius:4px; margin-top:0.4em;
}
.keybox{
  border:1px solid #d5dded; background:#f6f8fc;
  border-radius:6px; padding:0.5em 0.9em;
}
.widget{
  border:1px solid #d5dded; border-radius:8px; padding:0.7em 0.9em;
  background:#fbfcfe;
}
.posterior{ font-size:1.9em; color:var(--hs-accent2); font-weight:700; }
.reveal .columns{ align-items:center; }
</style>
```

# Roadmap {.center}

::: {.incremental}
- What statistical inference is trying to do
- The **maximum-likelihood estimate** (MLE)
- **Frequentist** inference: standard error, confidence interval, p-value
- **Bayesian** inference: prior, posterior, credible interval
:::

## Where we are

This is the **third and final part** of the review.

Parts I–II introduced probability, random variables, sampling, and estimators. Here we turn to **statistical inference**: how we use a sample to learn about unknown population parameters.

::: {.callout-ts}
The emphasis is on the intuition behind **likelihood**, **frequentist inference**, and **Bayesian inference**. We study one unknown probability with a loan-default example, then apply the same logic to a linear regression.
:::

# What is inference? {background-color="#1b3a6b"}

## Statistical inference: the goal

Suppose a probability model contains an unknown parameter $\theta$. We observe a sample
$$y=(y_1,\ldots,y_n)$$
and want to learn about $\theta$.

::: {.keybox}
- **Estimation** gives us plausible values for the unknown parameter.
- **Inference** tells us how uncertain we should be about those values.
:::

## Two frameworks {.smaller}

| Framework | Main idea |
|:---|:---|
| **Frequentist** | Treat the parameter as fixed but unknown and study how an estimator behaves across repeated samples |
| **Bayesian** | Represent uncertainty about the parameter with a probability distribution and update it using the observed data |

::: {.callout-ts}
For likelihood-based methods, both approaches begin with the **same ingredient**: a probability model for the data.
:::

# From a model to a likelihood {background-color="#1b3a6b"}

## A running example — loan defaults {.smaller}

A bank observes **50 comparable loans**, of which **8 default**. Let
$$D_i=\begin{cases}1,& \text{loan }i\text{ defaults},\\ 0,&\text{otherwise.}\end{cases}$$

Assume each loan has the same unknown default probability $p$:
$$D_i\overset{iid}{\sim}\operatorname{Bernoulli}(p),\qquad p\in(0,1).$$

If $k$ defaults occur among $n$ independent loans,
$$K=\sum_{i=1}^{n}D_i\sim\operatorname{Binomial}(n,p),\qquad n=50,\ k=8.$$

## Constructing the likelihood {.smaller}

For any proposed $p$, the probability of exactly $k=8$ defaults is
$$P(K=8\mid p)=\binom{50}{8}p^{8}(1-p)^{42}.$$

Once the data are observed, $k=8$ is **fixed**. We now read this as a function of the unknown $p$:
$$L(p;k)=\binom{50}{8}p^{8}(1-p)^{42}\ \propto\ p^{8}(1-p)^{42}.$$

Terms not involving $p$ do not change *which* value maximises it.

## What the likelihood is — and is not {.smaller}

More generally,
$$L(\theta;y)=f(y\mid\theta),$$
with the observed data $y$ held fixed and the likelihood viewed as a function of $\theta$.

::: {.callout-ts}
**Important.** A likelihood is **not** a probability distribution for the parameter. It need not integrate to one over $\theta$. It measures the **relative support** the data give to different parameter values.
:::

## R — the likelihood for $p$ {.smaller}

```{r}
#| fig-width: 7
#| fig-height: 3.8
n <- 50
k <- 8

p_grid  <- seq(0.001, 0.40, length.out = 500)
log_lik <- dbinom(k, size = n, prob = p_grid, log = TRUE)

# Scale the likelihood so its maximum is 1
rel_lik <- exp(log_lik - max(log_lik))
p_hat   <- k / n

plot(p_grid, rel_lik, type = "l", lwd = 2,
     xlab = "Default probability p", ylab = "Relative likelihood",
     main = "Likelihood for p: 8 defaults in 50 loans")
abline(v = p_hat, lty = 2, lwd = 2)
```

## The maximum-likelihood estimate {.smaller}

The likelihood peaks at
$$\hat p=\frac{8}{50}=0.16.$$

This is the **maximum-likelihood estimate (MLE)**.

::: {.keybox}
The data give greatest support to a default probability of about **16%**, though nearby values are also plausible. Inference asks *how much* uncertainty remains around this estimate.
:::

# Frequentist inference {background-color="#1b3a6b"}

## Uncertainty through repeated sampling {.smaller}

In the frequentist framework $p$ is a **fixed but unknown constant**. Before the sample is drawn the data are random, so the estimator $\hat p=K/n$ is random too.

$$\text{model}\longrightarrow\text{estimator}\longrightarrow\text{sampling distribution}\longrightarrow\text{SE, CI, test}.$$

## Three tools, three meanings {.smaller}

::: {.keybox}
- **Standard error** — the standard deviation of the estimator's sampling distribution; how much the estimate would vary from sample to sample.
- **95% confidence interval** — a *procedure* such that, under repeated sampling, about 95% of the intervals it produces contain the true parameter. The 95% refers to the procedure, not to this one interval.
- **p-value** — the probability, *under the null model*, of a test statistic at least as extreme as observed. It is **not** the probability that $H_0$ is true.
:::

For example, test $H_0:p=0.10$.

## R — frequentist inference for $p$ {.smaller}

```{r}
p_hat <- k / n

# Exact (Clopper-Pearson) test and interval
default_test <- binom.test(x = k, n = n, p = 0.10, conf.level = 0.95)

# Normal-approximation (Wald) SE and interval, for comparison
wald_se <- sqrt(p_hat * (1 - p_hat) / n)
wald_ci <- p_hat + c(-1, 1) * 1.96 * wald_se

cat("p_hat      =", round(p_hat, 4), "\n")
cat("Wald SE    =", round(wald_se, 4), "\n")
cat("Wald 95% CI=", round(wald_ci, 4), "\n\n")

default_test
```

## Reading the result {.smaller}

::: {.keybox}
- Point estimate $\hat p=0.16$.
- The exact **Clopper–Pearson** interval $[0.072,\,0.291]$ is *not* symmetric around $\hat p$ and differs slightly from the Wald interval $\hat p\pm1.96\,\widehat{\mathrm{SE}}$; the two agree more closely as $n$ grows.
- The p-value ($\approx 0.16$) asks how unusual 8 defaults in 50 would be if $p=0.10$. At the 5% level it does **not** reject $H_0:p=0.10$.
:::

> *How would our estimator behave if we repeatedly sampled loans from the same population?*

# Bayesian inference {background-color="#1b3a6b"}

## Posterior $\propto$ likelihood $\times$ prior {.smaller}

Bayesian inference starts from the same likelihood but adds a **prior** for $p$:
$$\pi(p\mid y)=\frac{L(p;y)\,\pi(p)}{\int L(p;y)\,\pi(p)\,dp},$$

$$\boxed{\ \text{Posterior}\ \propto\ \text{Likelihood}\ \times\ \text{Prior}\ }$$

- **Prior** $\pi(p)$ — belief about $p$ *before* the current data.
- **Likelihood** $L(p;y)$ — what the data say about different $p$.
- **Posterior** $\pi(p\mid y)$ — belief about $p$ *after* combining them.

## A simple prior {.smaller}

Suppose past experience suggests default rates around 10%. For illustration take
$$p\sim\operatorname{Beta}(2,18),\qquad E(p)=\frac{2}{2+18}=0.10.$$

::: {.callout-ts}
The Beta distribution is convenient for probabilities because **all of its mass lies between 0 and 1**.
:::

## Conjugacy — Beta prior, Beta posterior {.smaller}

For a Binomial likelihood and a $\operatorname{Beta}(a,b)$ prior, the posterior is again Beta:
$$p\mid y\ \sim\ \operatorname{Beta}(a+k,\ b+n-k).$$

With $a=2,\ b=18,\ k=8,\ n=50$:
$$p\mid y\ \sim\ \operatorname{Beta}(10,60),\qquad E(p\mid y)=\frac{10}{70}\approx 0.143.$$

::: {.keybox}
The sample alone suggests **16%**; the prior was centred at **10%**; the posterior combines both and lands **between** them.
:::

## R — Bayesian inference for $p$ {.smaller}

```{r}
#| fig-width: 7
#| fig-height: 3.4
a_prior <- 2; b_prior <- 18
a_post  <- a_prior + k
b_post  <- b_prior + n - k

posterior_mean    <- a_post / (a_post + b_post)
credible_interval <- qbeta(c(0.025, 0.975), a_post, b_post)
prob_above_10     <- 1 - pbeta(0.10, a_post, b_post)

round(c(posterior_mean       = posterior_mean,
        lower_95             = credible_interval[1],
        upper_95             = credible_interval[2],
        `P(p > 0.10 | data)` = prob_above_10), 4)

library(ggplot2)
p_grid <- seq(0.001, 0.40, length.out = 500)
plot_data <- data.frame(
  p = rep(p_grid, 2),
  density = c(dbeta(p_grid, a_prior, b_prior),
              dbeta(p_grid, a_post,  b_post)),
  distribution = rep(c("Prior: Beta(2,18)", "Posterior: Beta(10,60)"),
                     each = length(p_grid)))

ggplot(plot_data, aes(x = p, y = density, linetype = distribution)) +
  geom_line(linewidth = 1) +
  labs(title = "Prior and posterior for the default probability",
       x = "Default probability p", y = "Density", linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

## Try it: prior meets data {.updating-slide background-color="#fbfcfe"}

<style>
.reveal .updating-slide h2 { font-size: 1.45em; }
.reveal .updating-slide .columns { gap: 1.2em; align-items: flex-start; }
.reveal .updating-slide .controls-box{
  font-size: 0.52em; line-height: 1.15; padding: 0.7em;
  border: 1px solid #d9e2ef; border-radius: 12px; background: white;
}
.reveal .updating-slide input[type="range"]{ width: 150px; }
.reveal .updating-slide .posterior{ font-size: 2.0em; font-weight: 700; color:#c0392b; }
.reveal .updating-slide .note{ font-size: 0.55em; color:#555; margin-top: 0.8em; }
</style>

:::: {.columns}

::: {.column width="36%"}

::: {.controls-box}

```{ojs}
//| echo: false
viewof priorMean = Inputs.range([0.02, 0.50], {
  value: 0.10, step: 0.005, label: "Prior mean"
})

viewof priorStrength = Inputs.range([2, 200], {
  value: 20, step: 1, label: "Prior strength a+b"
})

viewof kObs = Inputs.range([0, 50], {
  value: 8, step: 1, label: "Defaults k (of n = 50)"
})
```

:::

:::

::: {.column width="64%"}

```{ojs}
//| echo: false
nObs = 50
aPrior = priorMean * priorStrength
bPrior = (1 - priorMean) * priorStrength
aPost  = aPrior + kObs
bPost  = bPrior + (nObs - kObs)
postMean = aPost / (aPost + bPost)

// Unnormalised Beta density; normalise numerically on the grid
xs = d3.range(0.001, 0.6, 0.002)
dx = 0.002
betaU = (x, a, b) => Math.exp((a - 1) * Math.log(x) + (b - 1) * Math.log(1 - x))
normalise = (a, b) => {
  const raw = xs.map(x => betaU(x, a, b));
  const area = d3.sum(raw) * dx;
  return xs.map((x, i) => ({x, y: raw[i] / area}));
}

priorCurve = normalise(aPrior, bPrior)
postCurve  = normalise(aPost, bPost)
likeCurve  = normalise(kObs + 1, nObs - kObs + 1)   // likelihood shape in p

// P(p > 0.10 | data) by numeric integration of the posterior
probAbove = d3.sum(postCurve.filter(d => d.x > 0.10).map(d => d.y)) * dx

Plot.plot({
  height: 300, marginLeft: 45,
  x: {label: "Default probability p", domain: [0, 0.55]},
  y: {label: "density", grid: true},
  marks: [
    Plot.line(priorCurve, {x: "x", y: "y", stroke: "#8aa0c2", strokeWidth: 2, strokeDasharray: "4 3"}),
    Plot.line(likeCurve,  {x: "x", y: "y", stroke: "#c0392b", strokeWidth: 1.5, strokeDasharray: "1 3"}),
    Plot.line(postCurve,  {x: "x", y: "y", stroke: "#1b3a6b", strokeWidth: 3}),
    Plot.ruleX([postMean], {stroke: "#1b3a6b", strokeWidth: 1}),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
//| echo: false
html`<div>
  <div class="muted">Posterior mean E(p | data)</div>
  <div class="posterior">${(postMean * 100).toFixed(1)}%</div>
  <div class="note">
    <b style="color:#8aa0c2">╌</b> prior &nbsp;·&nbsp;
    <b style="color:#c0392b">┈</b> likelihood &nbsp;·&nbsp;
    <b style="color:#1b3a6b">━</b> posterior &nbsp; Beta(${aPost.toFixed(1)}, ${bPost.toFixed(1)}) &nbsp;·&nbsp;
    P(p &gt; 0.10 | data) ≈ <b>${(probAbove * 100).toFixed(0)}%</b>
  </div>
</div>`
```

:::

::::

::: {.note}
A stronger prior pulls the posterior toward the prior mean; more data (or the same $k$ with the shape sharpened) lets the likelihood dominate.
:::

## Credible interval & direct probability {.smaller}

A 95% **credible interval** contains 95% of the posterior probability for $p$, given model, prior, and data. Unlike a confidence interval, we can ask direct questions such as $P(p>0.10\mid y)$.

::: {.callout-ts}
Here $P(p>0.10\mid y)\approx 0.85$, while the exact test does **not** reject $H_0:p=0.10$ (p $\approx 0.16$). No conflict — they answer different questions. The test: how surprising 8/50 would be *if* $p=0.10$. The posterior: how probable $p>0.10$ is *given* the data and prior.
:::

## Frequentist vs Bayesian {.smaller}

| Question | Frequentist | Bayesian |
|:---|:---|:---|
| Unknown parameter | fixed but unknown | uncertainty is a distribution |
| Data model | Binomial likelihood | same Binomial likelihood |
| Extra ingredient | repeated-sampling principle | prior distribution |
| Point summary | MLE $\hat p$ | posterior mean/median/mode |
| Uncertainty | sampling distribution | posterior distribution |
| Interval | confidence interval | credible interval |
| Probability statement | about repeated samples/procedures | about the parameter given model, prior, data |

Same observed data and likelihood — **different treatment of uncertainty**.

# From one parameter to regression {background-color="#1b3a6b"}

## The wage–education question {.smaller}

The loan example has one unknown, $p$. Econometric models usually have several. We ask:

> **How is hourly wage associated with years of education?**

Using the `wage1` dataset from the **wooldridge** package, let
$$Y_i=\log(\text{hourly wage}_i),\qquad X_i=\text{years of education}_i,$$
$$Y_i=\beta_0+\beta_1 X_i+\varepsilon_i.$$

::: {.callout-ts}
This is an **associational** regression. By itself it does not establish that education has a causal effect on wages.
:::

## R — the `wage1` data {.smaller}

```{r}
#| fig-width: 7
#| fig-height: 3.6
if (!requireNamespace("wooldridge", quietly = TRUE))
  stop("Please install the 'wooldridge' package before rendering.")

data("wage1", package = "wooldridge")
head(wage1[c("wage", "lwage", "educ")])

ggplot(wage1, aes(x = educ, y = lwage)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "Education and hourly wages",
       x = "Years of education", y = "Log hourly wage") +
  theme_minimal(base_size = 13)
```

## Building the regression likelihood {.smaller}

The equation specifies only the **conditional mean**:
$$E(Y_i\mid X_i=x_i)=\beta_0+\beta_1 x_i.$$

For a likelihood we add a model for the disturbances. Assume $\varepsilon_i\overset{iid}{\sim}N(0,\sigma^2)$, so
$$Y_i\mid X_i=x_i\ \sim\ N(\beta_0+\beta_1 x_i,\ \sigma^2),\qquad \theta=(\beta_0,\beta_1,\sigma).$$

$$f(y_i\mid x_i,\theta)=\frac{1}{\sqrt{2\pi\sigma^2}}\exp\!\left[-\frac{(y_i-\beta_0-\beta_1 x_i)^2}{2\sigma^2}\right].$$

## The log-likelihood {.smaller}

Assuming conditional independence,
$$L(\theta;y,X)=\prod_{i=1}^{n} f(y_i\mid x_i,\theta),$$

$$\ell(\theta)=-\frac{n}{2}\log(2\pi)-n\log\sigma-\frac{1}{2\sigma^2}\sum_{i=1}^{n}(y_i-\beta_0-\beta_1 x_i)^2.$$

## Why OLS appears {.smaller}

For a fixed $\sigma$, maximising the Gaussian log-likelihood over $\beta_0,\beta_1$ is the same as **minimising**
$$\sum_{i=1}^{n}(y_i-\beta_0-\beta_1 x_i)^2.$$

::: {.keybox}
Under normally distributed errors, the maximum-likelihood estimates of the regression coefficients are exactly the **ordinary least-squares** estimates.
:::

## R — maximise the likelihood directly {.smaller}

```{r}
y <- wage1$lwage
x <- wage1$educ

neg_loglik <- function(par, y, x) {
  beta0 <- par[1]; beta1 <- par[2]
  sigma <- exp(par[3])           # log(sigma) keeps sigma > 0
  mu    <- beta0 + beta1 * x
  -sum(dnorm(y, mean = mu, sd = sigma, log = TRUE))
}

mle_fit <- optim(par = c(0, 0, log(sd(y))), fn = neg_loglik,
                 y = y, x = x, method = "BFGS", hessian = TRUE)

mle_estimates <- c(beta0 = mle_fit$par[1],
                   beta1 = mle_fit$par[2],
                   sigma = exp(mle_fit$par[3]))
round(mle_estimates, 4)
```

The optimiser searches over $(\beta_0,\beta_1,\sigma)$ for the values that make the observed wages most compatible with the Gaussian model.

## R — MLE equals OLS {.smaller}

```{r}
fit <- lm(lwage ~ educ, data = wage1)
coef(fit)

# Coefficients agree
rbind(MLE = mle_estimates[c("beta0", "beta1")],
      OLS = coef(fit))

# Standard errors also agree: MLE (inverted Hessian) vs OLS
mle_se <- sqrt(diag(solve(mle_fit$hessian)))[1:2]
se_tab <- rbind(`MLE SE` = unname(mle_se),
                `OLS SE` = unname(summary(fit)$coefficients[, "Std. Error"]))
colnames(se_tab) <- c("beta0", "beta1")
round(se_tab, 4)
```

## R — the likelihood along the slope {.smaller}

```{r}
#| fig-width: 7
#| fig-height: 3.4
b0_hat  <- mle_estimates["beta0"]
sig_hat <- mle_estimates["sigma"]

b1_grid <- seq(mle_estimates["beta1"] - 0.03,
               mle_estimates["beta1"] + 0.03, length.out = 200)

profile_ll <- sapply(b1_grid, function(b1)
  sum(dnorm(y, mean = b0_hat + b1 * x, sd = sig_hat, log = TRUE)))

ll_data <- data.frame(beta1 = b1_grid, log_likelihood = profile_ll)

ggplot(ll_data, aes(x = beta1, y = log_likelihood)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = mle_estimates["beta1"], linetype = "dashed", linewidth = 1) +
  annotate("text", x = mle_estimates["beta1"], y = max(profile_ll),
           label = "MLE", vjust = -0.8) +
  labs(title = "Regression log-likelihood peaks at the MLE",
       x = expression(beta[1] ~ "(return to education)"), y = "Log-likelihood") +
  theme_minimal(base_size = 13)
```

# Frequentist regression inference {background-color="#1b3a6b"}

## Sampling distribution of the slope {.smaller}

$\beta_1$ is a **fixed but unknown parameter**. The estimate $\hat\beta_1$ comes from this sample; across repeated samples it would vary — its **sampling distribution**. From it we obtain:

- a **standard error** for $\hat\beta_1$;
- a **confidence interval** for $\beta_1$;
- a test of, say, $H_0:\beta_1=0$.

::: {.callout-ts}
In a log-wage regression the slope is approximately the **proportional** difference in wage associated with one more year of education.
:::

## R — summary and interval {.smaller}

```{r}
summary(fit)
confint(fit, "educ", level = 0.95)
```

## Interpretation {.smaller}

::: {.keybox}
$\hat\beta_1\approx 0.083$, SE $\approx 0.008$ (so $t\approx 11$, p-value $\approx 0$), 95% CI $\approx[0.068,\,0.098]$. One extra year of education is associated with about an **8% higher hourly wage** on average ($R^2\approx 0.19$, $n=526$).
:::

The SE measures sampling uncertainty of the slope; the CI is read through repeated sampling; the p-value evaluates $H_0:\beta_1=0$ under the model. Again — an association, not automatically a causal effect.

# Bayesian regression inference {background-color="#1b3a6b"}

## Same likelihood, add priors {.smaller}

The likelihood is unchanged:
$$Y_i\mid X_i,\beta_0,\beta_1,\sigma^2\ \sim\ N(\beta_0+\beta_1 X_i,\ \sigma^2).$$

Add weak priors, e.g.
$$\beta_0\sim N(0,10^2),\qquad \beta_1\sim N(0,1^2),$$
$$p(\beta_0,\beta_1,\sigma^2\mid y,X)\ \propto\ L(\beta_0,\beta_1,\sigma^2;y,X)\,p(\beta_0,\beta_1,\sigma^2).$$

::: {.callout-ts}
For realistic models the posterior is obtained **numerically**. We draw from it with `MCMCregress()`; the MCMC mechanics are not our focus — the draws let us summarise posterior uncertainty.
:::

## R — draw from the posterior {.smaller}

```{r}
if (!requireNamespace("MCMCpack", quietly = TRUE))
  stop("Please install the 'MCMCpack' package before rendering.")

bayes_fit <- MCMCpack::MCMCregress(
  lwage ~ educ, data = wage1,
  b0 = c(0, 0),                 # prior mean for (beta0, beta1)
  B0 = diag(c(0.01, 1)),        # prior precision: beta0 ~ N(0,10^2), beta1 ~ N(0,1^2)
  c0 = 0.001, d0 = 0.001,       # weak prior on the disturbance variance
  burnin = 2000, mcmc = 10000, seed = 821)

summary(bayes_fit)
```

## R — posterior for the slope {.smaller}

```{r}
#| fig-width: 7
#| fig-height: 3.2
beta1_post <- bayes_fit[, "educ"]

posterior_mean    <- mean(beta1_post)
posterior_sd      <- sd(beta1_post)
credible_interval <- quantile(beta1_post, c(0.025, 0.975))
prob_positive     <- mean(beta1_post > 0)

round(c(posterior_mean        = posterior_mean,
        posterior_sd          = posterior_sd,
        lower_95              = credible_interval[1],
        upper_95              = credible_interval[2],
        `P(beta1 > 0 | data)` = prob_positive), 4)

ggplot(data.frame(beta1_post = beta1_post), aes(x = beta1_post)) +
  geom_density(linewidth = 1) +
  geom_vline(xintercept = credible_interval, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = coef(fit)["educ"], linewidth = 1) +
  annotate("text", x = credible_interval[1], y = 0, label = "2.5%",  vjust = -1) +
  annotate("text", x = credible_interval[2], y = 0, label = "97.5%", vjust = -1) +
  annotate("text", x = coef(fit)["educ"],    y = 0, label = "OLS",   vjust = -1) +
  labs(title = "Posterior distribution of the education coefficient",
       x = expression(beta[1]), y = "Density") +
  theme_minimal(base_size = 13)
```

## Interpreting the posterior {.smaller}

::: {.keybox}
- **Posterior mean** — a point summary of $\beta_1$.
- **Posterior SD** — the size of posterior uncertainty.
- **95% credible interval** — 95% of the posterior probability for $\beta_1$.
- $P(\beta_1>0\mid y,X)$ — posterior probability the association is positive, given model and prior.
:::

With a sizeable sample and weak priors, Bayesian and frequentist **numbers** are often similar; their **interpretations** stay different.

## R — frequentist vs Bayesian, one table {.smaller}

```{r}
ols_est <- coef(fit)["educ"]
ols_se  <- summary(fit)$coefficients["educ", "Std. Error"]
ols_ci  <- confint(fit, "educ", level = 0.95)

comparison <- rbind(
  Frequentist = c(estimate = unname(ols_est), se = unname(ols_se),
                  lower_95 = unname(ols_ci[1]), upper_95 = unname(ols_ci[2])),
  Bayesian    = c(estimate = unname(posterior_mean), se = unname(posterior_sd),
                  lower_95 = unname(credible_interval[1]),
                  upper_95 = unname(credible_interval[2])))
round(comparison, 4)
```

## Side by side {.smaller}

| Question | Frequentist | Bayesian |
|:---|:---|:---|
| Parameter $\beta_1$ | fixed but unknown | posterior distribution |
| Probability model | Gaussian regression | same |
| Likelihood | $L(\beta_0,\beta_1,\sigma;y,X)$ | same |
| Extra ingredient | repeated-sampling principle | prior distribution |
| Point summary | OLS / MLE $\hat\beta_1$ | posterior mean/median/mode |
| Uncertainty | standard error | posterior SD |
| 95% interval | confidence interval | credible interval |
| $P(\beta_1>0\mid y)$ | not a parameter-probability | read from posterior draws |

**Same data. Same model. Same likelihood. Different treatment of uncertainty.**

# Where this leads {background-color="#1b3a6b"}

## Likelihood dominates here {.smaller}

On `wage1` both columns are almost identical — slope near **0.083**, 95% interval $\approx[0.068,\,0.098]$.

::: {.callout-ts}
With 526 observations and weak priors the **likelihood dominates**, so the posterior sits essentially on top of the OLS estimate. The two would diverge with a small sample or a strong prior — exactly where **Bayesian shrinkage** becomes valuable, as in the VARs later in the course.
:::

## Recap {.smaller}

- A **likelihood** measures the relative support the data give to parameter values — it is not itself a distribution over $\theta$.
- The **MLE** maximises it; under normal errors, MLE **is** OLS.
- **Frequentist** inference reads uncertainty through the sampling distribution: SE, confidence interval, p-value.
- **Bayesian** inference adds a prior and updates it: posterior mean, credible interval, direct probability statements.
- Same data and likelihood — the frameworks **differ in how they represent and interpret uncertainty**.

::: {.small .muted}
Notebook: `04-Review-stats2.ipynb` · R 4.6 · packages: ggplot2, wooldridge, MCMCpack. Interactive widget runs in the browser (Observable JS); the R figures are produced at render time.
:::

## {.center background-color="#1b3a6b"}

<h2 style="color:white">Questions?</h2>

<p style="color:#cdd8ea">HS 821 · Lecture 3 · Sunil Paul</p>
