---
title: "Univariate Time Series Models I"
subtitle: "HS 821 · Lecture 5 · Topics in Time Series Econometrics"
author: "Sunil Paul"
date: "2026-08-18"
format:
  revealjs:
    theme: default
    slide-number: c/t
    show-slide-number: all
    incremental: false
    smaller: false
    scrollable: true
    center-title-slide: true
    transition: slide
    background-transition: fade
    chalkboard: true
    code-line-numbers: false
    code-overflow: wrap
    html-math-method: mathjax
    fig-align: center
    footer: "HS 821 · Univariate Time Series Models I"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dev: "png"
    dpi: 150
    fig.width: 7
    fig.height: 4.2
---

```{=html}
<style>
:root{
  --hs-accent:#1b3a6b;
  --hs-accent2:#c0392b;
}
.reveal h1,.reveal h2,.reveal h3{ color:var(--hs-accent); font-weight:650; }
.reveal .slide-number{ background:transparent; color:#888; }
.reveal section img{ border:none; box-shadow:none; }
.reveal table{ font-size:0.80em; }
.reveal table th{ background:#eef2f8; }
.reveal pre code{ font-size:0.72em; line-height:1.25; }
.reveal .small{ font-size:0.78em; }
.reveal .muted{ color:#5a6472; }
.reveal blockquote{
  border-left:4px solid var(--hs-accent);
  background:#f6f8fc; padding:0.4em 0.9em; font-style:normal;
}
.callout-ts{
  border-left:5px solid var(--hs-accent2);
  background:#fdf1ef; padding:0.5em 0.9em; border-radius:4px; margin-top:0.4em;
}
.keybox{
  border:1px solid #d5dded; background:#f6f8fc;
  border-radius:6px; padding:0.5em 0.9em;
}
.widget{
  border:1px solid #d5dded; border-radius:8px; padding:0.7em 0.9em;
  background:#fbfcfe;
}
.posterior{ font-size:1.9em; color:var(--hs-accent2); font-weight:700; }
.reveal .columns{ align-items:center; }
</style>
```

```{r}
#| echo: false
#| include: false
# ------------------------------------------------------------------
# Setup: libraries and the plotting helpers used throughout the deck.
# (Hidden — these are the same helpers as the lecture notebook.)
# ------------------------------------------------------------------
suppressPackageStartupMessages({
  library(ggplot2)
  library(gridExtra)
})

# Time-series line plot; optional dashed reference line at the mean.
ts_plot <- function(x, mu = NULL) {
  stopifnot(is.numeric(x) || is.ts(x))
  time_idx <- if (is.ts(x)) as.numeric(time(x)) else seq_along(x)
  df <- data.frame(time = time_idx, value = as.numeric(x))
  p <- ggplot(df, aes(time, value)) +
    theme_classic() + geom_line() + labs(x = "Time", y = NULL)
  if (!is.null(mu)) p <- p + geom_hline(yintercept = mu, linetype = "dashed")
  p
}

# Theoretical ACF of an ARMA process, lags 1..L (lag 0 dropped).
acf_plot <- function(ar = numeric(0), ma = numeric(0), L = 16, title = NULL) {
  rho <- ARMAacf(ar = ar, ma = ma, lag.max = L)[-1]
  df  <- data.frame(lag = seq_len(L), acf = as.numeric(rho))
  ggplot(df, aes(lag, acf)) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_segment(aes(xend = lag, y = 0, yend = acf), linewidth = 1) +
    geom_point(size = 2) +
    scale_x_continuous(breaks = seq(0, L, by = 2)) +
    labs(x = "Lag", y = "ACF", title = title)
}

# Impulse response of an MA(q): psi = (1, theta_1, ..., theta_q, 0, ...).
irf_ma <- function(ma = numeric(0), h = 8, title = NULL) {
  psi <- c(1, ma, rep(0, max(0, h - length(ma))))[seq_len(h + 1)]
  df  <- data.frame(horizon = 0:h, response = psi)
  ggplot(df, aes(horizon, response)) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_segment(aes(xend = horizon, y = 0, yend = response)) +
    geom_point() +
    scale_x_continuous(breaks = 0:h) +
    labs(x = "Horizon after innovation", y = "Response", title = title)
}

# Running moment diagnostics used in the ergodicity simulations.
running_moments <- function(y) {
  Tn      <- length(y)
  lagprod <- y[-1] * y[-Tn]
  data.frame(
    t             = 1:Tn,
    mean          = cumsum(y)   / (1:Tn),
    second_moment = cumsum(y^2) / (1:Tn),
    lag1_product  = c(NA, cumsum(lagprod) / (1:(Tn - 1)))
  )
}
```

# Roadmap {.center}

::: {.incremental}
- **Stochastic processes** and the one-realization problem
- **Moments**: mean, autocovariance, autocorrelation
- **Stationarity** and, crucially, **ergodicity**
- **White noise** and the population-vs-sample ACF
- **Wold decomposition**, the lag operator, and linear processes
- **Moving-average** models: ACF, invertibility, impulse responses
:::

## Where we are

Earlier lectures treated observations largely as a **random sample**. Time-series data are different: **ordering matters**, observations may be dependent, and we usually observe only **one realization** of an underlying stochastic process.

::: {.callout-ts}
This lecture develops the **probabilistic properties** of univariate time series — the foundation for ARMA modelling and forecasting later in the course. Estimation, inference, and model selection come afterwards.
:::

# From samples to time series {background-color="#1b3a6b"}

## A stochastic process vs a realization {.smaller}

A **stochastic process** is a collection of random variables indexed by time,
$$\{Y_t:t\in\mathbb Z\}.$$

Before observation each $Y_t$ is random. Once observed over $t=1,\ldots,T$ we obtain **one realization**,
$$y_1,y_2,\ldots,y_T.$$

::: {.keybox}
A time series is **not** an unordered collection of observations. Its **temporal ordering and dependence structure** are part of the object being studied.
:::

> Shuffle a cross-section and little is lost. Shuffle a time series and its dependence — its time-series structure — is destroyed.

## Moments of a stochastic process {.smaller}

**Mean function:**
$$\mu_t=E(Y_t).$$

**Autocovariance** ($h$ periods apart):
$$\gamma(t,t-h)=\operatorname{Cov}(Y_t,Y_{t-h})=E[(Y_t-\mu_t)(Y_{t-h}-\mu_{t-h})].$$

For a **covariance-stationary** process these depend only on the lag $h$:
$$\gamma_h=\operatorname{Cov}(Y_t,Y_{t-h}),\qquad \rho_h=\frac{\gamma_h}{\gamma_0}.$$

::: {.keybox}
The sequence $\{\rho_h\}$ is the **autocorrelation function (ACF)**; its plot against $h$ is the **correlogram**.
:::

## Restrictions on an autocovariance function {.smaller}

For a covariance-stationary process,
$$\gamma_h=\gamma_{-h},\quad \rho_h=\rho_{-h},\quad \gamma_0=\operatorname{Var}(Y_t)\ge0,\quad \rho_0=1,\quad |\rho_h|\le1.$$

A deeper restriction: for any constants $a_1,\ldots,a_m$ and indices $t_1,\ldots,t_m$,
$$\sum_{i=1}^{m}\sum_{j=1}^{m}a_i a_j\,\gamma_{t_i-t_j}\ge0.$$

The autocovariance sequence must be **positive semidefinite**.

::: {.callout-ts}
**Question.** Is any symmetric sequence with $\rho_0=1$ and $|\rho_h|\le1$ a valid ACF? **No** — the autocorrelations must jointly define a positive-semidefinite covariance structure.
:::

# Stationarity {background-color="#1b3a6b"}

## Strict vs covariance stationarity {.smaller}

**Strict stationarity** — shifting time changes no finite-dimensional joint distribution:
$$(Y_{t_1},\ldots,Y_{t_k})\overset{d}{=}(Y_{t_1+h},\ldots,Y_{t_k+h}).$$

**Covariance (weak, second-order) stationarity:**

1. $E(Y_t)=\mu$ for all $t$;
2. $\operatorname{Var}(Y_t)=\gamma_0<\infty$ for all $t$;
3. $\operatorname{Cov}(Y_t,Y_{t-h})=\gamma_h$ depends only on the lag $h$.

::: {.keybox}
If second moments exist, strict $\Rightarrow$ covariance stationarity (not conversely). For a **Gaussian** process the mean and covariance pin down all distributions, so covariance stationarity $\Rightarrow$ strict.
:::

## Predict before you simulate {.smaller}

Consider four processes:

$$Y_t=\varepsilon_t \quad(\text{white noise}), \qquad Y_t=2+0.03t+\varepsilon_t \quad(\text{trend}),$$
$$Y_t=Y_{t-1}+\varepsilon_t \quad(\text{random walk}), \qquad Y_t=t\,\varepsilon_t \quad(\text{time-varying variance}).$$

For each, ask **before running**:

- Is the mean constant?
- Is the variance constant?
- If not stationary, **which condition fails?**

## R — simulate four processes {.smaller}

```{r}
#| fig-height: 5.2
set.seed(821)
T <- 250; t <- 1:T

wn    <- rnorm(T)
trend <- 2 + 0.03 * t + rnorm(T)
rw    <- cumsum(rnorm(T))
tvv   <- t * rnorm(T)

p1 <- ts_plot(wn,    mu = 0) + ggtitle("White noise")
p2 <- ts_plot(trend)         + ggtitle("Deterministic trend + noise")
p3 <- ts_plot(rw)            + ggtitle("Random walk")
p4 <- ts_plot(tvv,   mu = 0) + ggtitle("Time-varying variance")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Reading the graphs {.smaller}

::: {.keybox}
- **White noise** — constant mean and variance.
- **Deterministic trend** — violates the constant-mean condition.
- **Random walk** — $\operatorname{Var}(Y_t)=t\sigma^2$ (with $Y_0$ fixed): variance grows with time.
- **Time-varying variance** — $E(Y_t)=0$ but $\operatorname{Var}(Y_t)=t^2\sigma^2$: nonstationary through the variance.
:::

> **Stationarity is a property of the process, not a visual property of one observed path.**

# Ergodicity {background-color="#1b3a6b"}

## Stationarity is not enough {.smaller}

Stationarity keeps the probability law stable over time. **Ergodicity** connects two very different ways of averaging.

::: {.keybox}
- **Ensemble average** — average across **many parallel realizations** (histories) at a fixed date $t$. This is the expectation $E(Y_t)$.
- **Time average** — average **along one realization** over time: $\dfrac{1}{T}\sum_{t=1}^{T}Y_t$.
:::

A stationary process is **mean ergodic** when the two coincide as the sample grows:
$$\frac{1}{T}\sum_{t=1}^{T}Y_t \xrightarrow{p} E(Y_t).$$

The same idea extends to variances and autocovariances.

## Why it is the load-bearing assumption {.smaller}

We almost never get the ensemble — we get **one history**. One realized path of US GDP, one path of the S&P 500, one inflation series.

::: {.callout-ts}
Every time we estimate a mean, variance, or autocorrelation from that single path and treat it as a population quantity, we are **assuming ergodicity**. If it fails, more of the *same* history does not help.
:::

Two warnings the simulations make concrete:

- **stationarity $\not\Rightarrow$ ergodicity** — a stable law can hide a permanent, history-specific component;
- **serial dependence $\not\Rightarrow$ non-ergodicity** — a stationary AR(1) is dependent yet ergodic.

## A real-world picture: permanent vs transitory {.smaller}

The simulated example $Y_t = A + \varepsilon_t$ (with $A$ drawn once, $\varepsilon_t$ transitory) is the **permanent–transitory** decomposition of applied economics.

- $Y_t$ = a household's **log income**: permanent component $A$ (human capital, profession) plus a good/bad year $\varepsilon_t$ — the structure behind Friedman's *permanent-income hypothesis*.
- Or $Y_t$ = a **country's** growth: a permanent "type" $A$ (institutions, geography) plus year-to-year shocks.

::: {.keybox}
Follow **one** household for decades and its time-average income converges to **its own** $A$, not the population mean. Watching Argentina for two centuries reveals Argentina's $A$, not the world-average growth rate.
:::

Contrast a stationary AR(1): a shock's influence decays, $\gamma_h=\phi^h\to0$, so a long path *does* recover the ensemble moments.

## Simulation 1 — iid white noise {.smaller}

$$Y_t\overset{iid}{\sim}N(0,1),\qquad E(Y_t)=0,\quad E(Y_t^2)=1,\quad \gamma_1=0.$$

We track three running time averages at once — the mean, the second moment, and the lag-1 product.

> **Predict:** as the realization lengthens, what should each of the three approach?

## R — ergodic white noise {.smaller}

```{r}
#| fig-height: 5.6
set.seed(821)
T <- 5000; y <- rnorm(T)
d <- running_moments(y)

p1 <- ggplot(d, aes(t, mean)) + geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") + theme_classic() +
  labs(title = "Running mean", x = "T", y = expression(frac(1, T) * sum(Y[t])))

p2 <- ggplot(d, aes(t, second_moment)) + geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed") + theme_classic() +
  labs(title = "Running second moment", x = "T", y = expression(frac(1, T) * sum(Y[t]^2)))

p3 <- ggplot(d, aes(t, lag1_product)) + geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") + theme_classic() +
  labs(title = "Running lag-1 product", x = "T",
       y = expression(frac(1, T - 1) * sum(Y[t] * Y[t - 1])))

grid.arrange(p1, p2, p3, ncol = 1)
```

## What you should see {.smaller}

As $T$ increases,
$$\frac{1}{T}\sum Y_t \xrightarrow{p} 0,\qquad \frac{1}{T}\sum Y_t^2 \xrightarrow{p} 1,\qquad \frac{1}{T-1}\sum_{t=2}^{T}Y_t Y_{t-1} \xrightarrow{p} 0.$$

::: {.keybox}
The **path** never becomes smooth. What stabilizes are the **time averages of functions of the process**.
:::

## Simulation 2 — dependence $\ne$ non-ergodicity {.smaller}

Stationary AR(1): $\;Y_t=\phi Y_{t-1}+\varepsilon_t,\ |\phi|<1.$

Set $\phi=0.8$ and $\operatorname{Var}(\varepsilon_t)=1-\phi^2$, so
$$E(Y_t)=0,\qquad \operatorname{Var}(Y_t)=1,\qquad \gamma_1=0.8.$$

Serially dependent — yet **ergodic**.

> **Predict:** should the running mean converge to zero? What should the running lag-1 product approach?

## R — stationary, dependent, yet ergodic {.smaller}

```{r}
#| fig-height: 5.6
set.seed(822)
T <- 5000; phi <- 0.8
innovation_sd <- sqrt(1 - phi^2)          # normalises Var(Y_t) to 1

y_ar <- as.numeric(arima.sim(model = list(ar = phi), n = T, sd = innovation_sd))
d    <- running_moments(y_ar)

p1 <- ggplot(d, aes(t, mean)) + geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") + theme_classic() +
  labs(title = "AR(1): running mean", x = "T", y = NULL)

p2 <- ggplot(d, aes(t, second_moment)) + geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed") + theme_classic() +
  labs(title = "AR(1): running second moment", x = "T", y = NULL)

p3 <- ggplot(d, aes(t, lag1_product)) + geom_line() +
  geom_hline(yintercept = phi, linetype = "dashed") + theme_classic() +
  labs(title = "AR(1): running lag-1 product",
       subtitle = expression(gamma[1] == 0.8), x = "T", y = NULL)

grid.arrange(p1, p2, p3, ncol = 1)
```

## AR(1): the limits {.smaller}

$$\frac{1}{T}\sum Y_t \xrightarrow{p} 0,\qquad \frac{1}{T}\sum Y_t^2 \xrightarrow{p} 1,\qquad \frac{1}{T-1}\sum_{t=2}^{T}Y_t Y_{t-1} \xrightarrow{p} \gamma_1 = 0.8.$$

> **Ergodicity does not require independence.** Strongly serially correlated data can still reveal the stationary moments through a long enough realization.

## Simulation 3 — stationary but non-ergodic {.smaller}

$$Y_t=A+\varepsilon_t,\qquad A\sim N(0,1),\quad \varepsilon_t\overset{iid}{\sim}N(0,1),$$

with $A$ drawn **once** and held fixed. Unconditionally it is stationary:
$$E(Y_t)=0,\quad \operatorname{Var}(Y_t)=2,\quad \gamma_h=\operatorname{Var}(A)=1\ (h>0).$$

But within one realization $A$ never averages out. As $T\to\infty$,
$$\tfrac{1}{T}\sum Y_t \to A,\qquad \tfrac{1}{T}\sum Y_t^2 \to A^2+1,\qquad \tfrac{1}{T-1}\sum Y_t Y_{t-1} \to A^2,$$

not the ensemble values $0,\ 2,\ 1$.

## R — a permanent random component {.smaller}

```{r}
#| fig-height: 6.0
set.seed(823)
T <- 5000
A <- rnorm(1); epsilon <- rnorm(T)
y_ne <- A + epsilon
d <- running_moments(y_ne)

p0 <- ts_plot(y_ne) + geom_hline(yintercept = A, linetype = "dashed") +
  ggtitle("One realization: Y_t = A + epsilon_t")

p1 <- ggplot(d, aes(t, mean)) + geom_line() +
  geom_hline(yintercept = 0, linetype = "dotted") +
  geom_hline(yintercept = A, linetype = "dashed") + theme_classic() +
  labs(title = "Running mean",
       subtitle = "Dotted: ensemble 0; dashed: limit A", x = "T", y = NULL)

p2 <- ggplot(d, aes(t, second_moment)) + geom_line() +
  geom_hline(yintercept = 2, linetype = "dotted") +
  geom_hline(yintercept = A^2 + 1, linetype = "dashed") + theme_classic() +
  labs(title = "Running second moment",
       subtitle = "Dotted: ensemble 2; dashed: limit A^2 + 1", x = "T", y = NULL)

p3 <- ggplot(d, aes(t, lag1_product)) + geom_line() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_hline(yintercept = A^2, linetype = "dashed") + theme_classic() +
  labs(title = "Running lag-1 product",
       subtitle = "Dotted: ensemble 1; dashed: limit A^2", x = "T", y = NULL)

grid.arrange(p0, p1, p2, p3, ncol = 1)
cat("Realization-specific A =", round(A, 4), "\n")
```

## Why stationarity does not imply ergodicity {.smaller}

$Y_t=A+\varepsilon_t$ is stationary, but the common component $A$ creates dependence that **never dies out**:
$$\gamma_h=1\qquad\text{for every }h>0.$$

By contrast, for a stationary AR(1),
$$\gamma_h\propto\phi^h\to0\qquad(|\phi|<1).$$

::: {.keybox}
The issue is not *whether* observations are dependent, but whether the dependence is **persistent enough** to stop time averages from revealing ensemble moments.
:::

## Many realizations make it visible {.smaller}

Generate many realizations of increasing length and compare their **time averages**.

- Ergodic process → time averages concentrate around the ensemble mean.
- $Y_t=A+\varepsilon_t$ → transitory noise averages out, but $A$ remains.

> Predict the two panels before running.

## R — time averages across realizations {.smaller}

```{r}
#| fig-height: 5.0
set.seed(824)
T_values <- c(25, 100, 500, 2000); R <- 300

results <- do.call(rbind, lapply(T_values, function(Ti) {
  wn_means         <- replicate(R, mean(rnorm(Ti)))
  nonergodic_means <- replicate(R, { A_i <- rnorm(1); mean(A_i + rnorm(Ti)) })
  rbind(
    data.frame(T = factor(Ti, levels = T_values),
               process = "Ergodic white noise",         time_mean = wn_means),
    data.frame(T = factor(Ti, levels = T_values),
               process = "Stationary but non-ergodic",  time_mean = nonergodic_means))
}))

ggplot(results, aes(x = T, y = time_mean)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ process, ncol = 1) + theme_classic() +
  labs(title = "Time averages across many realizations",
       x = "Length of realization T", y = "Time average")
```

## Interpretation {.smaller}

::: {.keybox}
For white noise the distribution of time averages **collapses** toward zero as $T$ grows. For $Y_t=A+\varepsilon_t$ each realization converges to its own $A$, so the distribution **does not** collapse to the ensemble mean.
:::

> Ergodicity is what lets a long history substitute, in an appropriate sense, for repeated draws from the ensemble.

## Why ergodicity changes decisions {.smaller}

A repeated gamble on wealth $W_t$ — each period, independently,
$$W_t = 1.5\,W_{t-1}\ (\text{prob }0.5),\qquad W_t = 0.6\,W_{t-1}\ (\text{prob }0.5).$$

**Ensemble view** (across many parallel gamblers):
$$E\!\left[\tfrac{W_t}{W_{t-1}}\right]=0.5(1.5)+0.5(0.6)=1.05 \;\Rightarrow\; E[W_t]=(1.05)^t W_0.$$

**Time view** (one gambler's path):
$$\tfrac{1}{t}\log\tfrac{W_t}{W_0}\xrightarrow{p} 0.5\log 1.5+0.5\log 0.6 \approx -0.053.$$

::: {.callout-ts}
Expected wealth grows ~5%/period, but the **typical path shrinks ~5%/period** toward zero. This is **ergodicity economics** (Ole Peters); it is why geometric returns, volatility drag, and the Kelly criterion matter.
:::

## R — the ensemble vs the path you live {.smaller}

```{r}
#| fig-height: 4.0
set.seed(826)
n_paths <- 5000; Tn <- 100

multipliers <- matrix(sample(c(1.5, 0.6), n_paths * Tn, replace = TRUE), nrow = n_paths)
wealth <- cbind(1, t(apply(multipliers, 1, cumprod)))   # W_0 = 1

t_idx         <- 0:Tn
ensemble_mean <- colMeans(wealth)
median_path   <- apply(wealth, 2, median)
geo_rate      <- exp(0.5 * log(1.5) + 0.5 * log(0.6))    # ~ 0.9487

plot_df <- rbind(
  data.frame(t = t_idx, wealth = ensemble_mean, series = "Ensemble average E[W_t]"),
  data.frame(t = t_idx, wealth = median_path,   series = "Typical (median) path"))

ggplot(plot_df, aes(t, wealth, colour = series)) +
  geom_line(linewidth = 1) + scale_y_log10() + theme_classic() +
  theme(legend.position = "top", legend.title = element_blank()) +
  labs(title = "Ensemble average rises; the typical path collapses",
       x = "Period", y = "Wealth (log scale)")

cat(sprintf("Ensemble growth/period : %.3f\n", 1.05))
cat(sprintf("Time-average growth    : %.3f  (geometric mean)\n", geo_rate))
```

## Summary: stationarity and ergodicity {.smaller}

| Process | Stationary? | Dependence? | Mean ergodic? | Long-run covariance |
|:---|:---:|:---:|:---:|:---|
| iid white noise | Yes | No | Yes | $\gamma_h=0$ for $h>0$ |
| stationary AR(1), $|\phi|<1$ | Yes | Yes | Yes | $\gamma_h\to0$ |
| $Y_t=A+\varepsilon_t$ | Yes | Yes | No | $\gamma_h=1$ for all $h>0$ |
| random walk | No | Yes | — | variance grows with $t$ |

$$\boxed{\text{stationary}\not\Rightarrow\text{ergodic}},\qquad \boxed{\text{dependence}\not\Rightarrow\text{non-ergodic}}.$$

# White noise {background-color="#1b3a6b"}

## White noise {.smaller}

A **white-noise process** $\{\varepsilon_t\}$ satisfies
$$E(\varepsilon_t)=0,\quad \operatorname{Var}(\varepsilon_t)=\sigma^2,\quad \operatorname{Cov}(\varepsilon_t,\varepsilon_{t-h})=0\ (h\neq0),$$
so its theoretical ACF is $\rho_0=1,\ \rho_h=0\ (h\neq0)$.

::: {.keybox}
White noise means **serially uncorrelated** — not necessarily independent.

- **iid white noise:** independent and identically distributed.
- **Gaussian white noise:** a *jointly* Gaussian white noise; here zero autocovariance **does** imply independence.
- **martingale difference:** $E(\varepsilon_t\mid\mathcal F_{t-1})=0$.
:::

## Population ACF vs sample ACF {.smaller}

The **population ACF** is a property of the process; the **sample ACF** is an estimate from one finite realization.

For white noise, $\rho_h=0\ (h>0)$ — but a finite sample almost never gives sample autocorrelations that are exactly zero.

> **Predict:** with $T=100$ vs $T=1000$, which sample ACF lies more tightly around zero?

## R — population vs sample ACF {.smaller}

```{r}
#| fig-height: 5.0
set.seed(825)
sample_acf_df <- function(n, max_lag = 20) {
  a <- acf(rnorm(n), plot = FALSE, lag.max = max_lag)
  data.frame(lag = as.numeric(a$lag[-1]), acf = as.numeric(a$acf[-1]),
             n = paste0("T = ", n), bound = 1.96 / sqrt(n))
}
acf_samples <- rbind(sample_acf_df(100), sample_acf_df(1000))
bounds <- unique(acf_samples[c("n", "bound")])

ggplot(acf_samples, aes(lag, acf)) +
  geom_hline(yintercept = 0) +
  geom_segment(aes(xend = lag, y = 0, yend = acf)) + geom_point() +
  geom_hline(data = bounds, aes(yintercept =  bound), linetype = "dashed") +
  geom_hline(data = bounds, aes(yintercept = -bound), linetype = "dashed") +
  facet_wrap(~ n, ncol = 1) + theme_classic() +
  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
  labs(title = "Sample ACF of white noise",
       subtitle = "Dashed: approximate ±1.96/sqrt(T) bounds", x = "Lag", y = "Sample ACF")
```

## Interpretation {.smaller}

::: {.keybox}
The population ACF is zero at every positive lag, yet the sample ACF fluctuates around zero because it comes from **one realization**. As $T$ grows the fluctuations shrink — where the distinction between **process** and **realization** becomes operational.
:::

# A stationary process and a random walk {background-color="#1b3a6b"}

## Same shocks, different processes {.smaller}

Let $X_t=+1$ or $-1$ with probability $0.5$ each (independent). Then
$$E(X_t)=0,\quad \operatorname{Var}(X_t)=1,\quad \gamma_h=0\ (h\neq0),$$
so $\{X_t\}$ is covariance stationary.

Accumulate the **same** shocks:
$$Y_t=Y_{t-1}+X_t,\ Y_0=0 \;\Rightarrow\; Y_t=\sum_{i=1}^{t}X_i,\quad \operatorname{Var}(Y_t)=t.$$

> The shocks are stationary; their accumulation is not.

## R — same shocks, different properties {.smaller}

```{r}
#| fig-height: 4.4
set.seed(100)
n <- 250
X <- sample(c(-1, 1), n, replace = TRUE)
Y <- cumsum(X)

p1 <- ts_plot(X, mu = 0) + ggtitle("Stationary shocks: X_t")
p2 <- ts_plot(Y)         + ggtitle("Accumulated shocks: random walk Y_t")
grid.arrange(p1, p2, ncol = 1)
```

## Differencing changes the process {.smaller}

For the random walk $Y_t=Y_{t-1}+\varepsilon_t$, the level is nonstationary because $\operatorname{Var}(Y_t)=t\sigma^2$. But
$$\Delta Y_t=Y_t-Y_{t-1}=\varepsilon_t,$$
which is stationary white noise.

::: {.callout-ts}
The distinction between the properties of a **level** and its **difference** becomes central when we study **unit roots and integrated processes**.
:::

# Linear representations {background-color="#1b3a6b"}

## Wold decomposition {.smaller}

For a covariance-stationary, **purely nondeterministic** process,
$$Y_t-\mu=\varepsilon_t+\psi_1\varepsilon_{t-1}+\psi_2\varepsilon_{t-2}+\cdots=\sum_{j=0}^{\infty}\psi_j\varepsilon_{t-j},$$
with $\psi_0=1$ and $\sum_{j=0}^{\infty}\psi_j^2<\infty$.

$\varepsilon_t$ is the **innovation** — the one-step-ahead **linear projection error**. Innovations are serially uncorrelated by construction; they need not be iid or Gaussian.

::: {.callout-ts}
A Wold innovation is a **statistical** innovation. It is not automatically a **causal structural shock**, and it is defined relative to *linear* prediction — a nonlinear forecaster may use information a linear projection cannot.
:::

## Moments of a linear process {.smaller}

For $Y_t=\mu+\sum_{j=0}^{\infty}\psi_j\varepsilon_{t-j},\ \varepsilon_t\sim WN(0,\sigma^2)$:
$$E(Y_t)=\mu,\qquad \gamma_0=\sigma^2\sum_{j=0}^{\infty}\psi_j^2,$$
$$\gamma_h=\sigma^2\sum_{j=0}^{\infty}\psi_j\psi_{j+h}\qquad(h\ge0).$$

::: {.keybox}
This links the **dynamic response coefficients** $\{\psi_j\}$ to the **dependence structure** $\{\gamma_h\}$.
:::

## Lag operator {.smaller}

The **lag operator** $L$ shifts back one period:
$$LY_t=Y_{t-1},\qquad L^jY_t=Y_{t-j}.$$

For example,
$$Y_t=\mu+\varepsilon_t+\theta_1\varepsilon_{t-1}+\theta_2\varepsilon_{t-2}$$
becomes
$$Y_t-\mu=(1+\theta_1L+\theta_2L^2)\varepsilon_t=\theta(L)\varepsilon_t.$$

Lag-polynomial notation makes AR, MA, and ARMA properties far easier to express.

## Familiar linear-process families {.smaller}

With white-noise innovations $\varepsilon_t$:

$$\textbf{AR}(p):\quad Y_t=c+\phi_1Y_{t-1}+\cdots+\phi_pY_{t-p}+\varepsilon_t.$$
$$\textbf{MA}(q):\quad Y_t=\mu+\varepsilon_t+\theta_1\varepsilon_{t-1}+\cdots+\theta_q\varepsilon_{t-q}.$$
$$\textbf{ARMA}(p,q):\quad Y_t=c+\sum_{i=1}^{p}\phi_iY_{t-i}+\varepsilon_t+\sum_{j=1}^{q}\theta_j\varepsilon_{t-j}.$$

::: {.keybox}
This lecture concentrates on **MA processes**: a finite MA is automatically stationary when the innovations are white noise with finite variance. AR and ARMA follow later.
:::

# Moving-average processes {background-color="#1b3a6b"}

## MA(1): moments and ACF {.smaller}

$$Y_t=\mu+\varepsilon_t+\theta\varepsilon_{t-1},\qquad \varepsilon_t\sim WN(0,\sigma^2).$$

$$E(Y_t)=\mu,\qquad \gamma_h=\begin{cases}(1+\theta^2)\sigma^2,&h=0,\\ \theta\sigma^2,&|h|=1,\\ 0,&|h|>1.\end{cases}$$

$$\rho_1=\frac{\theta}{1+\theta^2},\qquad \rho_h=0\ (|h|>1).$$

::: {.keybox}
- ACF **cuts off after lag 1**; the sign of $\rho_1$ is the sign of $\theta$.
- Since $\left|\frac{\theta}{1+\theta^2}\right|\le\frac12$, an MA(1) can **never** have $|\rho_1|>0.5$.
:::

## Predict before running {.smaller}

Compare
$$Y_t=2+\varepsilon_t+0.8\varepsilon_{t-1}\qquad\text{and}\qquad Y_t=2+\varepsilon_t-0.8\varepsilon_{t-1}.$$

Before looking at the graphs:

1. What sign should $\rho_1$ have in each?
2. What is the numerical value of $\rho_1$?
3. At which lag must the theoretical ACF become exactly zero?
4. Must the two simulated paths look obviously different?

## R — two MA(1) processes {.smaller}

```{r}
#| fig-height: 5.0
set.seed(123)
n <- 250
ma_pos <- 2 + arima.sim(model = list(ma =  0.8), n = n)
ma_neg <- 2 + arima.sim(model = list(ma = -0.8), n = n)

p1 <- ts_plot(ma_pos, mu = 2) + ggtitle("MA(1): theta = 0.8")
p2 <- acf_plot(ma =  0.8, L = 12, title = "Theoretical ACF: theta = 0.8")
p3 <- ts_plot(ma_neg, mu = 2) + ggtitle("MA(1): theta = -0.8")
p4 <- acf_plot(ma = -0.8, L = 12, title = "Theoretical ACF: theta = -0.8")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## Reading the graphs {.smaller}

For $\theta=0.8$, $\ \rho_1=\dfrac{0.8}{1+0.8^2}\approx0.488$; for $\theta=-0.8$, $\ \rho_1\approx-0.488$. In both, $\rho_h=0$ for $|h|>1$.

::: {.keybox}
The left plots are particular **realizations**; the right plots are the **theoretical ACFs** of the underlying processes. Same $|\rho_1|$, opposite sign.
:::

## Try it: MA(1) explorer {.updating-slide background-color="#fbfcfe"}

<style>
.reveal .updating-slide h2 { font-size: 1.45em; }
.reveal .updating-slide .columns { gap: 1.2em; align-items: flex-start; }
.reveal .updating-slide .controls-box{
  font-size: 0.60em; line-height: 1.2; padding: 0.7em;
  border: 1px solid #d9e2ef; border-radius: 12px; background: white;
}
.reveal .updating-slide input[type="range"]{ width: 160px; }
.reveal .updating-slide .posterior{ font-size: 2.0em; font-weight: 700; color:#c0392b; }
.reveal .updating-slide .note{ font-size: 0.60em; color:#555; margin-top: 0.8em; }
</style>

:::: {.columns}

::: {.column width="36%"}

::: {.controls-box}

```{ojs}
//| echo: false
viewof theta = Inputs.range([-1.5, 1.5], {
  value: 0.8, step: 0.05, label: "MA(1) coefficient θ"
})
```

:::

:::

::: {.column width="64%"}

```{ojs}
//| echo: false
rho1 = theta / (1 + theta * theta)
Hmax = 8
acfData = d3.range(0, Hmax + 1).map(h => ({
  lag: h, rho: h === 0 ? 1 : (h === 1 ? rho1 : 0)
}))

Plot.plot({
  height: 300, marginLeft: 45,
  x: {label: "lag h", domain: [-0.5, Hmax + 0.5]},
  y: {label: "ρ(h)", domain: [-0.6, 1.05], grid: true},
  marks: [
    Plot.ruleY([0]),
    Plot.ruleX(acfData, {x: "lag", y1: 0, y2: "rho", stroke: "#1b3a6b", strokeWidth: 3}),
    Plot.dot(acfData, {x: "lag", y: "rho", fill: "#1b3a6b", r: 4})
  ]
})
```

```{ojs}
//| echo: false
html`<div>
  <div class="muted">Lag-1 autocorrelation &nbsp; ρ₁ = θ / (1 + θ²)</div>
  <div class="posterior">${rho1.toFixed(3)}</div>
  <div class="note">
    |θ| = ${Math.abs(theta).toFixed(2)} ⇒
    <b>${Math.abs(theta) < 1 ? "invertible" : (Math.abs(theta) > 1 ? "not invertible" : "unit root")}</b>
    &nbsp;·&nbsp; ACF cuts off after lag 1
    &nbsp;·&nbsp; note |ρ₁| ≤ 0.5, reached at θ = ±1
  </div>
</div>`
```

:::

::::

::: {.note}
Move $\theta$: the sign of $\rho_1$ follows the sign of $\theta$, its magnitude never exceeds $0.5$, and $\theta$ vs $1/\theta$ give the **same** $\rho_1$ — the invertibility ambiguity.
:::

## MA($q$) {.smaller}

$$Y_t=\mu+\sum_{j=0}^{q}\theta_j\varepsilon_{t-j},\qquad \theta_0=1,\quad \varepsilon_t\sim WN(0,\sigma^2).$$

For $h=0,1,\ldots,q$,
$$\gamma_h=\sigma^2\sum_{j=0}^{q-h}\theta_j\theta_{j+h},\qquad \gamma_h=0\ (h>q),$$
so the theoretical ACF **cuts off after lag $q$**.

::: {.keybox}
The cutoff is specific to linear dependence through the innovations: once $h>q$, $Y_t$ and $Y_{t-h}$ share **no common innovations**.
:::

## MA(2): work from the common shocks {.smaller}

$$Y_{1t}=2+\varepsilon_t+1.0\,\varepsilon_{t-1}+0.7\,\varepsilon_{t-2},\qquad Y_{2t}=2+\varepsilon_t-1.2\,\varepsilon_{t-1}+0.25\,\varepsilon_{t-2}.$$

Before running, use $\gamma_h=\sigma^2\sum_{j=0}^{q-h}\theta_j\theta_{j+h}$ to answer:

1. At which lag must each ACF become zero?
2. For $Y_{1t}$, what signs do $\gamma_1,\gamma_2$ take?
3. For $Y_{2t}$, can $\gamma_1$ and $\gamma_2$ differ in sign?

## R — two MA(2) processes {.smaller}

```{r}
#| fig-height: 5.0
set.seed(123)
n <- 250
ma21 <- c(1.0, 0.7); ma22 <- c(-1.2, 0.25)

s21 <- 2 + arima.sim(model = list(ma = ma21), n = n)
s22 <- 2 + arima.sim(model = list(ma = ma22), n = n)

p1 <- ts_plot(s21, mu = 2) + ggtitle("MA(2): theta = (1.0, 0.7)")
p2 <- acf_plot(ma = ma21, L = 12, title = "Theoretical ACF")
p3 <- ts_plot(s22, mu = 2) + ggtitle("MA(2): theta = (-1.2, 0.25)")
p4 <- acf_plot(ma = ma22, L = 12, title = "Theoretical ACF")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## MA(2): the autocovariances {.smaller}

Using $\gamma_h=\sigma^2\sum_{j=0}^{q-h}\theta_j\theta_{j+h}$ with $\theta_0=1$:

For $Y_{1t}$:
$$\gamma_1=\sigma^2(\theta_0\theta_1+\theta_1\theta_2)=1.7\,\sigma^2,\qquad \gamma_2=\sigma^2\theta_0\theta_2=0.7\,\sigma^2.$$

For $Y_{2t}$:
$$\gamma_1=\sigma^2[-1.2+(-1.2)(0.25)]=-1.5\,\sigma^2,\qquad \gamma_2=0.25\,\sigma^2.$$

::: {.keybox}
An MA(2) can display autocorrelations of **different signs** at lags 1 and 2; both ACFs are exactly zero for $h>2$.
:::

# Invertibility {background-color="#1b3a6b"}

## Invertibility {.smaller}

Every finite MA is covariance stationary, but different parameterizations can share the same autocovariance structure. For an MA(1),
$$\rho_1=\frac{\theta}{1+\theta^2},$$
so $\theta$ and $1/\theta$ give the same $\rho_1$ (e.g. $\theta=0.5$ and $\theta=2$ both give $\rho_1=0.4$).

**Invertibility** picks the representation in which the innovation is recoverable from current and past observations:
$$\varepsilon_t=(1+\theta L)^{-1}(Y_t-\mu)=(1-\theta L+\theta^2L^2-\cdots)(Y_t-\mu),$$
which converges when $|\theta|<1$.

::: {.callout-ts}
Invertibility is about the **uniqueness and recoverability** of the innovation representation — **not** a stationarity condition.
:::

## R — equivalent parameterizations {.smaller}

```{r}
equivalent_ma1 <- data.frame(theta = c(0.5, 2.0), sigma2 = c(1.0, 0.25))
equivalent_ma1$gamma0 <- equivalent_ma1$sigma2 * (1 + equivalent_ma1$theta^2)
equivalent_ma1$gamma1 <- equivalent_ma1$sigma2 * equivalent_ma1$theta
equivalent_ma1$rho1   <- equivalent_ma1$gamma1 / equivalent_ma1$gamma0
equivalent_ma1
```

Same $\gamma_0,\gamma_1,\rho_1$ from two different $(\theta,\sigma^2)$ pairs — only $|\theta|<1$ is invertible.

## Root condition {.smaller}

For $Y_t-\mu=(1+\theta L)\varepsilon_t$, invertibility requires the root of $1+\theta z=0$ to lie **outside the unit circle**. The root is $z=-1/\theta$, so for an MA(1), $|\theta|<1$.

More generally, an MA($q$) is invertible when every root of
$$1+\theta_1z+\cdots+\theta_qz^q=0$$
lies outside the unit circle.

> A single coefficient larger than one does **not** by itself decide invertibility for MA(2)+ — the condition is on the **roots of the whole lag polynomial**.

## R — check invertibility {.smaller}

```{r}
ma_roots <- function(theta) {
  roots <- polyroot(c(1, theta))
  data.frame(root = roots, modulus = Mod(roots),
             outside_unit_circle = Mod(roots) > 1)
}
ma_roots(ma21)
ma_roots(ma22)
```

Both MA(2) examples have all roots outside the unit circle — both invertible.

# Impulse responses {background-color="#1b3a6b"}

## Wold coefficients as dynamic responses {.smaller}

In $Y_t-\mu=\sum_{j=0}^{\infty}\psi_j\varepsilon_{t-j}$, the coefficient $\psi_j$ is the **linear response at horizon $j$** to a one-unit innovation today.

For an MA($q$) the response is exactly zero after horizon $q$; for an infinite-order MA it may decay gradually.

::: {.callout-ts}
These are responses to **statistical innovations**. A causal interpretation requires **additional identification assumptions**.
:::

> **Predict** for $Y_t=\varepsilon_t+0.8\varepsilon_{t-1}$: the response immediately, one period later, two periods later?

## R — dynamic responses {.smaller}

```{r}
#| fig-height: 3.6
p1 <- irf_ma(0.8,         h = 8, title = "Dynamic response: MA(1)")
p2 <- irf_ma(c(1.0, 0.7), h = 8, title = "Dynamic response: MA(2)")
grid.arrange(p1, p2, ncol = 2)
```

The MA(1) response is $\psi=(1,\,0.8,\,0,\ldots)$; the MA(2) is $\psi=(1,\,1.0,\,0.7,\,0,\ldots)$ — zero beyond the order $q$.

# Recap {background-color="#1b3a6b"}

## Key takeaways {.smaller}

::: {.keybox}
- A **stochastic process** vs one observed **realization** — most properties are about the process.
- **Covariance stationarity** stabilizes the law; **ergodicity** equates time and ensemble averages. Stationary $\not\Rightarrow$ ergodic; dependence $\not\Rightarrow$ non-ergodic.
- **White noise** is uncorrelated, not necessarily independent; the sample ACF is only an estimate of the population ACF.
- The **Wold decomposition** writes any covariance-stationary process as a linear filter of innovations; the lag operator streamlines the algebra.
- **MA($q$)** ACFs cut off after lag $q$; **invertibility** (roots outside the unit circle) selects the recoverable innovation representation.
- Wold coefficients $\psi_j$ are **impulse responses** to statistical innovations.
:::

## Exercises {.smaller}

1. **What property fails?** For $Y_t=5+\varepsilon_t$, $\ 5+0.1t+\varepsilon_t$, $\ t\varepsilon_t$, $\ Y_{t-1}+\varepsilon_t$: is each covariance stationary, and which condition fails?
2. **Stationarity vs ergodicity.** $Y_t=A,\ A\sim N(0,1)$: why stationary, and why does a long record not reveal the ensemble mean?
3. **MA(1) ACF.** For $Y_t=3+\varepsilon_t+0.6\varepsilon_{t-1},\ \varepsilon_t\sim WN(0,4)$: find $E(Y_t),\gamma_0,\gamma_1,\rho_1$ and the cutoff lag.
4. **MA(2) from common innovations.** For $Y_t=\varepsilon_t+0.7\varepsilon_{t-1}-0.3\varepsilon_{t-2}$: derive $\gamma_0,\gamma_1,\gamma_2$ (predict signs first).
5. **Valid ACF?** Why is $|\rho_h|\le1$ insufficient for $\rho_0=1,\rho_1=0.9,\rho_2=-0.9$? What else is required?
6. **Invertibility.** Show $\theta=0.5$ and $\theta=2$ (with suitable variances) give the same MA(1) autocovariance; which is invertible, and what does that mean for the innovations?

## Suggested reading {.smaller}

- Hamilton, J. D. (1994), *Time Series Analysis*, Ch. 3.
- Brockwell, P. J. and Davis, R. A., *Introduction to Time Series and Forecasting* — stationary processes and ARMA.
- Enders, W., *Applied Econometric Time Series* — stationary stochastic processes and ARMA.
- Hyndman, R. J. and Athanasopoulos, G., *Forecasting: Principles and Practice* — time-series structure and ARIMA, for an intuitive computational treatment.

> Next lecture: **autoregressive processes** — stationarity conditions, characteristic roots, persistence, and the contrasting ACF/PACF behaviour of AR and MA.

::: {.small .muted}
Source notebook: `05-Univariate-Time-Series-modelsI.ipynb` · R 4.6 · packages: ggplot2, gridExtra. R figures are produced at render time; the MA(1) explorer runs in the browser (Observable JS).
:::
